---
title: "Gene Utility Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene Utility Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# All Functions

To see the fucntions currently available, do:

micromicon_functions()


# Gene Utility Functions

The `micromicon` package provides a comprehensive set of utility functions for working with genes in `genome_entity` objects. These functions make it easy to extract sequences, translate proteins, map positions, and validate variants.

## Overview

The gene utility functions include:

### Genomic Features

- **`get_features()`** - List all annotated features as a data.frame
- **`search_features()`** - filter features by type, name pattern, or coordinates
- **`get_roi_features()`** - features overlapping a genomic region
- **`gene_info()`** - full details for a single gene

### Sequences

- **`get_gene_dna()`** - Extract DNA sequence for a gene in 5'-to-3' canonical orientation
- **`get_gene_aa()`** - Get translated amino acid sequence for a CDS
- **`get_roi_dna()`** - Extract DNA for an arbitrary genomic region
- **`sequences()`** - List all scaffold (contig/chromosome/plasmid) sequences

### Coordinate Mapping

- **`map_genomic_to_cds()`** - Convert genomic position to CDS position
- **`map_cds_to_genomic()`** - Convert CDS position to genomic position
- **`extract_by_name()`** - Extract sequence and features by gene name
- **`extract_by_coords()`** - Extract sequence and features by coordinates
- **`gene_info()`** - Get comprehensive gene metadata

### Genome Metadata

- **`get_genome_metadata()`** - Get organism name, length, topology, etc.
- **`get_genome_seqnames()`** - List contig / chromosome names

### Variants & mutations 

- **`predict_mutations()`** - Get a tidy mutation table from a genome diff
- **`compute_effects()`** - Look at molecular consequences (aa change, consequence class)
- **`validate_variant_in_gene()`** - Check whether a variant falls within a gene

## Setup

```{r setup, eval=FALSE}
library(micromicon)

# Load a genome
entity <- read_genome("path/to/genome.gbk")
# OR
entity <- read_genome(gff = "path/to/genome.gff3", fasta = "path/to/genome.fasta")
```

## Getting Gene Sequences

### DNA Sequence

Extract the DNA sequence for a gene in 5'-to-3' orientation (automatically handles reverse complement for minus strand genes):

```{r get-dna, eval=FALSE}
# By feature index
dna <- get_gene_dna(entity, 1)

# By locus tag
dna <- get_gene_dna(entity, "b0001")

# By gene name
dna <- get_gene_dna(entity, "dnaA")

# Get as Biostrings DNAString (requires Biostrings package)
dna_string <- get_gene_dna(entity, "dnaA", format = "DNAString")
```

### Amino Acid Sequence

Get the translated amino acid sequence for a CDS:

```{r get-aa, eval=FALSE}
# Use default genetic code (11 - bacterial/archaeal)
aa <- get_gene_aa(entity, "dnaA")

# Use standard genetic code (1)
aa <- get_gene_aa(entity, "dnaA", genetic_code = "1")

# The function automatically uses the transl_table attribute if present
# in the feature annotation
```

### Region of Interest (ROI)

Extract DNA for any arbitrary genomic region:

```{r get-roi, eval=FALSE}
# Get 100 bp region from position 1000-1099 on plus strand
roi <- get_roi_dna(entity, chrom = "chromosome", start = 1000, end = 1099, strand = "+")

# Get region on minus strand (automatically reverse complements)
roi_minus <- get_roi_dna(entity, chrom = "chromosome", start = 1000, end = 1099, strand = "-")
```

## Translation

`micromicon` has numerous inner functions as well, such as `translate_dna()` to multiple reading frames and genetic codes. NOTE that this function issue uses the STANDARD genetic code translation, which does NOT account for  :

```{r translate, eval=FALSE}
dna <- "ATGAAATTTCCCGGGTTTAAATAG"

# Translate in frame 1 (default)
aa1 <- micromicon:::translate_dna(dna, frame = 1, genetic_code = "11") <- Note ":::" to access additional functions

# Translate in frame 2 (skip first base)
aa2 <- micromicon:::translate_dna(dna, frame = 2, genetic_code = "11")

# Translate in frame 3 (skip first two bases)
aa3 <- micromicon:::translate_dna(dna, frame = 3, genetic_code = "11")

# Use standard genetic code
aa_std <- translate_dna(dna, frame = 1, genetic_code = "1")
```

### Supported Genetic Codes

Currently supported NCBI genetic codes:
- `"1"` - Standard (universal) genetic code
- `"11"` - Bacterial, Archaeal, and Plant Plastid Code (default)

## Position Mapping

### Genomic to CDS Position

Convert a genomic coordinate to a CDS-relative position (accounts for strand):

```{r map-to-cds, eval=FALSE}
# For a gene on the plus strand
genomic_pos <- 12345
cds_pos <- map_genomic_to_cds(entity, "dnaA", genomic_pos)
# If gene starts at 12000, cds_pos will be 346

# For a gene on the minus strand
# The function automatically accounts for reverse orientation
cds_pos <- map_genomic_to_cds(entity, "some_minus_gene", genomic_pos)
```

### CDS to Genomic Position

Convert a CDS position back to genomic coordinates:

```{r map-to-genomic, eval=FALSE}
# Convert CDS position 100 to genomic coordinate
cds_pos <- 100
genomic_pos <- map_cds_to_genomic(entity, "dnaA", cds_pos)

# This is the inverse of map_genomic_to_cds()
```

## Gene Information

Get comprehensive metadata about a gene:

```{r gene-info, eval=FALSE}
info <- gene_info(entity, "dnaA")

# Returns a list with:
# - chrom: Chromosome/contig name
# - start: Start position (1-based)
# - end: End position (1-based)
# - strand: Strand ("+" or "-")
# - type: Feature type (e.g., "CDS")
# - length: Gene length in bp
# - cds_ranges: List of CDS coordinate ranges
# - frame: Reading frame offset (0, 1, or 2)
# - gene: Gene name
# - locus_tag: Locus tag
# - product: Product description

cat(sprintf("Gene %s is located at %s:%d-%d\n",
            info$gene, info$chrom, info$start, info$end))
cat(sprintf("Length: %d bp on %s strand\n",
            info$length, info$strand))
```

## Variant Validation

Validate that a genomic variant is correctly positioned and has the expected reference base:

```{r validate-variant, eval=FALSE}
# Validate a SNP
result <- validate_variant_in_gene(
  entity,
  gene = "dnaA",
  genomic_pos = 12345,
  ref_base = "A"
)

if (result) {
  cat("Valid variant!\n")
} else {
  cat(sprintf("Invalid: %s\n", attr(result, "message")))
}

# Common validation failures:
# - Position outside gene boundaries
# - Reference base mismatch
# - Position on wrong chromosome
```

## Gene Identifier Types

All gene utility functions support multiple ways to identify genes:

1. **By integer index**: `get_gene_dna(entity, 1)` - Uses the first feature
2. **By locus tag**: `get_gene_dna(entity, "b0001")` - Searches locus_tag field
3. **By gene name**: `get_gene_dna(entity, "dnaA")` - Searches gene field
4. **By ID**: `get_gene_dna(entity, "gene-001")` - Searches ID field
5. **By feature row**: `get_gene_dna(entity, features(entity)[1,])` - Pass data.frame row

## Example Workflow

Here's a complete example of analyzing a gene:

```{r workflow, eval=FALSE}
library(micromicon)

# Load genome
entity <- read_genome("ecoli.gbk")

# Find gene of interest
gene_name <- "dnaA"

# Get gene information
info <- gene_info(entity, gene_name)
cat(sprintf("Analyzing gene: %s\n", info$gene))
cat(sprintf("Location: %s:%d-%d (%s)\n",
            info$chrom, info$start, info$end, info$strand))
cat(sprintf("Length: %d bp, Frame: %d\n", info$length, info$frame))

# Get sequences
dna <- get_gene_dna(entity, gene_name)
aa <- get_gene_aa(entity, gene_name)
cat(sprintf("DNA: %d bp\n", nchar(dna)))
cat(sprintf("Protein: %d aa\n", nchar(aa)))

# Check if there's a premature stop codon
if (grepl("\\*", substr(aa, 1, nchar(aa) - 1))) {
  cat("WARNING: Premature stop codon detected!\n")
}

# Validate a known variant
variant_pos <- info$start + 50
ref_base <- substr(get_roi_dna(entity, info$chrom, variant_pos, variant_pos), 1, 1)
is_valid <- validate_variant_in_gene(entity, gene_name, variant_pos, ref_base)
cat(sprintf("Variant at position %d: %s\n",
            variant_pos, ifelse(is_valid, "VALID", "INVALID")))

# Calculate CDS position for variant
cds_pos <- map_genomic_to_cds(entity, gene_name, variant_pos)
cat(sprintf("Variant is at CDS position %d\n", cds_pos))

# Determine which codon is affected
codon_num <- ceiling(cds_pos / 3)
codon_pos <- ((cds_pos - 1) %% 3) + 1
cat(sprintf("Affects codon %d, position %d\n", codon_num, codon_pos))
```

## Working with Variation Analysis Mode

All gene utility functions work seamlessly with `genome_entity_gd` objects (breseq output):

```{r variation-mode, eval=FALSE}
# Load reference genome
entity <- read_genome("reference.gbk")

# Parse breseq output
gd <- parse_gd_annotated("annotated.gd", entity)

# Get predicted mutations
mutations <- predict_mutations(gd)

# Use gene utilities on gd objects (inheritance)
gene_dna <- get_gene_dna(gd, "mutS")
gene_aa <- get_gene_aa(gd, "mutS")

# Validate variants from mutation table
for (i in seq_len(nrow(mutations))) {
  mut <- mutations[i, ]

  if (mut$type == "SNP" && !is.na(mut$gene)) {
    # Validate the mutation
    is_valid <- validate_variant_in_gene(
      gd,
      gene = mut$gene,
      position = mut$position,
      ref_base = mut$ref
    )

    cat(sprintf("SNP in %s: %s\n",
                mut$gene,
                ifelse(is_valid, "VALID", "INVALID")))
  }
}
```

For more details on working with breseq data, see `vignette("variation-analysis")`.

## Performance Notes

- All functions validate inputs and provide informative error messages
- DNA sequences are stored as character vectors for efficiency
- Position mappings use 1-based coordinates (standard in biology)
- Functions automatically handle strand orientation (reverse complement)
- Genetic code translation is optimized for bacterial genomes (code 11)

## See Also

- `features()` - Access genome features
- `sequences()` - Access genome sequences
- `search_features()` - Search for features by pattern
- `extract_by_coords()` - Extract sequences by coordinates
- `parse_gd_annotated()` - Parse breseq genome diffs
