---
title: "Variation Analysis with micromicon"
author: "William E. Ackerman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variation Analysis with micromicon}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE  # Set to TRUE when example data is available
)
```

# Introduction

`micromicon` supports two complementary modes for working with microbial genomes:

1. **Genome Navigation Mode** (`genome_entity`) - For exploring reference genomes
2. **Variation Analysis Mode** (`genome_entity_gd`) - For tracking mutations from breseq

This vignette focuses on **Variation Analysis Mode**, which integrates mutation data from the [breseq pipeline](https://barricklab.org/twiki/bin/view/Lab/ToolsBacterialGenomeResequencing) with reference genome navigation.

## Key Principle

`genome_entity_gd` is a **subclass** of `genome_entity`:

- All `genome_entity` functions work on `genome_entity_gd` (inheritance)
- `genome_entity_gd`-specific functions (like `predict_mutations()`) **don't work** on plain `genome_entity`

```{r class-hierarchy}
library(micromicon)

# Read reference genome
entity <- read_genome(fasta = "reference.fasta", gff = "reference.gff3")
class(entity)
# [1] "genome_entity"

# Parse breseq output
gd <- parse_gd_annotated("annotated.gd", entity)
class(gd)
# [1] "genome_entity_gd" "genome_entity"  # Subclass!
```

# Reading breseq Output

## Prerequisites

Before using `parse_gd_annotated()`, you need:

1. A **reference genome** (GenBank or GFF3+FASTA)
2. An **annotated.gd** file from breseq

```{r read-reference}
# Read reference genome
entity <- read_genome(fasta = "NC_000913.fasta",
                     gff = "NC_000913.gff3")
```

## Parsing annotated.gd

The `parse_gd_annotated()` function parses breseq's `annotated.gd` output and links it to your reference genome:

```{r parse-gd}
# Parse breseq output
gd <- parse_gd_annotated(
  gd_path = "output/annotated.gd",
  entity = entity
)

# Summary of mutations
summary(gd)
```

The resulting `genome_entity_gd` object contains:

- All reference genome data (sequences, features, metadata)
- Parsed mutation events (SNPs, insertions, deletions, etc.)
- Provenance information (breseq version, run parameters)
- Reference manifest (which reference sequences were used)

# Working with Mutation Data

## Predicted Mutations Table

Get a summary table of predicted mutations (reproduces breseq's "Predicted Mutations" table):

```{r predicted-mutations}
# Get mutations table
mutations <- predict_mutations(gd)

# View structure
head(mutations)

# Filter by type
snps <- mutations[mutations$type == "SNP", ]
indels <- mutations[mutations$type %in% c("INS", "DEL"), ]
```

The table includes:

- **seq_id** - Sequence identifier (chromosome/plasmid)
- **position** - Genomic position
- **type** - Mutation type (SNP, INS, DEL, SUB, MOB, etc.)
- **ref** - Reference allele
- **new** - Alternate allele
- **annotation** - Gene/feature annotation
- **description** - Detailed mutation description

## Mutation Types

breseq reports various mutation types:

- **SNP** - Single nucleotide polymorphism
- **DEL** - Deletion
- **INS** - Insertion
- **SUB** - Substitution (multi-bp)
- **MOB** - Mobile element insertion
- **AMP** - Amplification
- **CON** - Conversion
- **INV** - Inversion

```{r mutation-types}
# Count by type
table(mutations$type)

# Get large deletions
large_dels <- mutations[mutations$type == "DEL" &
                       nchar(mutations$ref) > 100, ]

# Get mobile element insertions
mob_events <- mutations[mutations$type == "MOB", ]
```

# Genome Navigation with Mutations

All standard genome navigation functions work on `genome_entity_gd` objects:

## Searching Features

```{r search-features}
# Search for genes (same as genome_entity)
ampC <- search_features(gd, pattern = "ampC", type = "CDS")
acrB <- search_features(gd, pattern = "acrB", type = "CDS")
```

## Extracting Sequences

```{r extract-sequences}
# Get gene sequences
protein <- get_gene_aa(gd, gene = "ampC")
dna <- get_gene_dna(gd, "acrB_1")

# Get regions with flanking
upstream <- get_feature_fasta(gd, "acrB_1",
                             flank_upstream = 5000,
                             flank_downstream = 2000)

# Get regions of interest
roi_seq <- get_roi_dna(gd, seqname = "1",
                      start = 1834322, end = 1837471)
```

## Coordinate Mapping

```{r coordinate-mapping}
# Map genomic position to CDS position
cds_pos <- map_genomic_to_cds(gd, gene = "dnaA",
                              genomic_pos = 3176824)

# Map CDS position to genomic position
genomic_pos <- map_cds_to_genomic(gd, gene = "dnaA",
                                  cds_pos = 3)
```

# Workflow Example

## Complete Analysis Pipeline

Here's a complete workflow for analyzing breseq results:

```{r complete-workflow}
library(micromicon)

# 1. Read reference genome
entity <- read_genome(fasta = "reference.fasta",
                     gff = "reference.gff3")

# 2. Parse breseq output
gd <- parse_gd_annotated("output/annotated.gd", entity)

# 3. Get mutation summary
mutations <- predict_mutations(gd)

# 4. Identify genes with mutations
mutated_genes <- unique(mutations$gene[!is.na(mutations$gene)])
cat(sprintf("Found mutations in %d genes\n", length(mutated_genes)))

# 5. Extract sequences for mutated genes
for (gene in mutated_genes[1:5]) {  # First 5 genes
  # Get reference sequence
  ref_seq <- get_gene_dna(gd, gene)

  # Get protein sequence
  protein <- get_gene_aa(gd, gene)

  # Find mutations in this gene
  gene_muts <- mutations[mutations$gene == gene, ]

  cat(sprintf("\n%s: %d mutation(s)\n", gene, nrow(gene_muts)))
  print(gene_muts[, c("position", "type", "description")])
}

# 6. Export sequences for downstream analysis
write_fasta(gd, "reference_sequences.fasta")
write_gff3(gd, "reference_features.gff3")
```

## Filtering Mutations

```{r filter-mutations}
# Get high-confidence SNPs (frequency > 0.9)
high_conf_snps <- mutations[mutations$type == "SNP" &
                            mutations$frequency > 0.9, ]

# Get coding mutations
coding_muts <- mutations[!is.na(mutations$gene), ]

# Get intergenic mutations
intergenic_muts <- mutations[is.na(mutations$gene), ]

# Get nonsynonymous mutations
nonsyn <- mutations[grepl("nonsynonymous", mutations$description,
                         ignore.case = TRUE), ]
```

## Comparing Multiple Samples

```{r multiple-samples}
# Parse multiple breseq outputs
samples <- c("sample1", "sample2", "sample3")
gd_list <- list()

for (sample in samples) {
  gd_path <- file.path(sample, "output", "annotated.gd")
  gd_list[[sample]] <- parse_gd_annotated(gd_path, entity)
}

# Get mutations for each sample
mut_list <- lapply(gd_list, predict_mutations)

# Find shared mutations
shared_genes <- Reduce(intersect,
                      lapply(mut_list, function(m) unique(m$gene)))

cat(sprintf("Found %d genes mutated in all samples\n",
           length(shared_genes)))
```

# Advanced Topics

## Accessing Raw Event Data

The `genome_entity_gd` object contains detailed event information:

```{r raw-events}
# Access raw events
str(gd$events, max.level = 1)

# Events are parsed from breseq output
length(gd$events)

# Each event contains detailed fields
names(gd$events[[1]])
```

## Provenance Information

Track breseq run parameters:

```{r provenance}
# breseq version and parameters
gd$provenance

# Reference manifest
gd$reference_manifest
```

## Integration with Other Tools

Export mutation data for other tools:

```{r export}
# Export mutations as CSV
mutations <- predict_mutations(gd)
write.csv(mutations, "mutations.csv", row.names = FALSE)

# Export VCF (future feature)
# write_vcf(gd, "mutations.vcf")
```

# Summary

**Variation Analysis Mode** (`genome_entity_gd`) provides:

- Seamless integration of breseq mutation data with reference genomes
- All genome navigation functions work unchanged (inheritance)
- Mutation-specific functions like `predict_mutations()`
- Provenance tracking for reproducibility

**Next steps**:

- See `vignette("micromicon-intro")` for general genome navigation
- See `vignette("gene-utilities")` for gene-level operations
- Check `?parse_gd_annotated` for detailed documentation

# References

- [breseq documentation](https://barricklab.org/twiki/bin/view/Lab/ToolsBacterialGenomeResequencing)
- [breseq GitHub](https://github.com/barricklab/breseq)
- Deatherage, D. E., & Barrick, J. E. (2014). Identification of mutations in laboratory-evolved microbes from next-generation sequencing data using breseq. *Methods in Molecular Biology*, 1151, 165-188.
